---
title: "Create Variables"
author: "Ellie Cox"
date: "2022-09-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages

```{r load}
library(dplyr)
library(lubridate)
library(tidyverse)
library(data.table)
library(ggplot2)
```

## Load Violation and Water System Data

```{r loaddata}
####################
### FIPS Data
####################
## Use fips codes from census data
#FIPS <- read.csv("/Users/elizabeth/Documents/Capstone Paper/FIPS_Codes.csv", skip = 4)

####################
### Violations Data
####################

setwd("/Users/elizabeth/Documents/Capstone Paper/Violations")
violation_files <- list.files(pattern="*.csv")

Violations <- rbindlist(lapply(violation_files, fread), fill = TRUE)

# Remove duplicated rows (I don't think the EPA did this, I think I might have just downloaded some data points multiple times)
Violations <- Violations[!duplicated(Violations)]

##### Change date format and add a date variable with just year
## COMPL_PER_BEGIN_DATE is the start date a facility has a violation
Violations$COMPL_PER_BEGIN_DATE <- lubridate::dmy(Violations$COMPL_PER_BEGIN_DATE)
Violations$year <- year(Violations$COMPL_PER_BEGIN_DATE)

####################
### All Public Water Systems
####################

multmerge = function(mypath){
        filenames = list.files(path = mypath, full.names = TRUE)
        rbindlist(lapply(filenames,function(i) tryCatch(fread(i,colClasses = list(character = c(12,35))),
                                                        error = function(e) {
                                                          cat("\nError reading in file:",i,"\t") #Identifies problem files by name
                                                          message(e) #Prints error message without stopping loop
                                                          list(ERROR=i) #Adds a placeholder column so rbindlist will execute
                                                          })), #End of tryCatch and lapply
                  fill = TRUE) #rbindlist arguments
        } #End of function

PWS = multmerge("/Users/elizabeth/Documents/Capstone Paper/Water System Data")

# Remove duplicated rows (I don't think the EPA did this, I think I might have just downloaded some data points multiple times)
PWS <- PWS[!duplicated(PWS)]
PWS <- PWS[!(PWS$PWS_NAME == "" | is.na(PWS$PWS_NAME)),]

##### Change date format and add a date variable with just year
PWS$PWS_DEACTIVATION_DATE <- lubridate::dmy(PWS$PWS_DEACTIVATION_DATE)
PWS$year <- year(PWS$PWS_DEACTIVATION_DATE)

####################
### Facility Registration Data
####################

frs <- read.csv("/Users/elizabeth/Documents/Capstone Paper/national_single/NATIONAL_SINGLE.CSV")
```

## Merge with FRS data to attach PWS to FIPS code

```{r altmerge}
## Create data frame to merge with that only has the fields I need
frs_cols <- data.frame(frs$PRIMARY_NAME, frs$FIPS_CODE)
colnames(frs_cols)[1] <- "PWS_NAME"; colnames(frs_cols)[2] <- "FIPS_CODE"
# Remove blank or NA entries
frs_cols <- frs_cols[!(frs_cols$PWS_NAME == "" | is.na(frs_cols$PWS_NAME)),]

# Perform left_join based on the public water system name
pws_frs <- merge(PWS, frs_cols, by = c("PWS_NAME"))

# Clean this up a bit
pws_frs <- pws_frs[!duplicated(pws_frs)]; pws_frs <- pws_frs[!(pws_frs$FIPS_CODE == ""| is.na(pws_frs$FIPS_CODE))]
```

Now merge with violations data

```{r violation_FIPS}
# Create data fram to merge with only the fields I want
PWSID_FIPS <- data.frame(pws_frs$PWSID, pws_frs$FIPS_CODE)
colnames(PWSID_FIPS)[1] <- "PWSID"; colnames(PWSID_FIPS)[2] <- "FIPS_CODE"

# Perform merge by PWSID
violations_fips <- merge(Violations, PWSID_FIPS, by = "PWSID")
```

Below calculates count variables that will be used to construct the final data matrix. 
First: the number of violating facilities in each county by year
Second: The number of facilities that have deactivated in a county in a particular year
Third: The total number of facilities in a county

```{r violations_county_year}
violation_facility <- data.frame(violations_fips$year, violations_fips$FIPS_CODE, violations_fips$PWSID)
violation_facility <- violation_facility[!duplicated(violation_facility),]
colnames(violation_facility)[1] <- "year"; colnames(violation_facility)[2] <- "FIPS_CODE"
colnames(violation_facility)[3] <- "PWSID"
 
violations_county_year <- violation_facility %>% 
  filter(year <= 2022) %>% 
  #filter(VIOLATION_CATEGORY_CODE == "MCL") %>% 
  group_by(year, FIPS_CODE) %>%
  count() 
```

```{r deactivations_county_year}
deactivations_county_year <- pws_frs %>% count(year, FIPS_CODE) %>% group_by(year, FIPS_CODE) %>% count()
```

```{r active_pws}
all_pws <- pws_frs %>% count(FIPS_CODE)
```

## Start constructing variables for regression

Each row is the number of violations for county x in year y. Make a matrix of the necessary dimension and se the "counts" calculated above to fill the observations that are not zero. 

Data covers years from 1974 to 2022
There are 5634 water systems total in this time period (active and deactive)

```{r numviolatingfacilities}
# Create empty dataframe to populate
Data <- data.frame(matrix(0, nrow = 49 * 5634, ncol = 5))
colnames(Data)[1] <- "year"; colnames(Data)[2] <- "FIPS"
colnames(Data)[3] <- "NumViolatingFacilities"; colnames(Data)[4] <- "AllFacilities"
colnames(Data) [5] <- "CumulativeDeactivatedFacilities"

# Populate year and FIPS entries
years <- 1974:2022
Data$year <- sort(rep(years, 5634))
codes <- unique(all_pws$FIPS_CODE) %>% sort()
Data$FIPS <- rep(codes, 49)

######################################
# Populate NumViolatingFacilities
Data_violations <- left_join(Data, violations_county_year, by = c("year" = "year", "FIPS" = "FIPS_CODE"))
Data_violations$n[is.na(Data_violations$n)] <- 0
Data$NumViolatingFacilities <- Data_violations$n

########################
# Populate ActiveFacilities
Data$AllFacilities <- rep(all_pws$n, 49)

## I'm sure there's a faster way of doing this, but this is what we're doing.
d1974 <- deactivations_county_year %>% filter(year == 1974)
d1975 <- deactivations_county_year %>% filter(year <= 1975) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1975)
d1976 <- deactivations_county_year %>% filter(year <= 1976) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1976)
d1977 <- deactivations_county_year %>% filter(year <= 1977) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1977)
d1978 <- deactivations_county_year %>% filter(year <= 1978) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1978)
d1979 <- deactivations_county_year %>% filter(year <= 1979) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1979)
d1980 <- deactivations_county_year %>% filter(year <= 1980) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1980)
d1981 <- deactivations_county_year %>% filter(year <= 1981) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1981)
d1982 <- deactivations_county_year %>% filter(year <= 1982) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1982)
d1983 <- deactivations_county_year %>% filter(year <= 1983) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1983)

d1984 <- deactivations_county_year %>% filter(year <= 1984) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1984)
d1985 <- deactivations_county_year %>% filter(year <= 1985) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1985)
d1986 <- deactivations_county_year %>% filter(year <= 1986) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1986)
d1987 <- deactivations_county_year %>% filter(year <= 1987) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1987)
d1988 <- deactivations_county_year %>% filter(year <= 1988) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1988)
d1989 <- deactivations_county_year %>% filter(year <= 1989) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1989)
d1990 <- deactivations_county_year %>% filter(year <= 1990) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1990)
d1991 <- deactivations_county_year %>% filter(year <= 1991) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1991)
d1992 <- deactivations_county_year %>% filter(year <= 1992) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1992)
d1993 <- deactivations_county_year %>% filter(year <= 1993) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1993)

d1994 <- deactivations_county_year %>% filter(year <= 1994) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1994)
d1995 <- deactivations_county_year %>% filter(year <= 1995) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1995)
d1996 <- deactivations_county_year %>% filter(year <= 1996) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1996)
d1997 <- deactivations_county_year %>% filter(year <= 1997) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1997)
d1998 <- deactivations_county_year %>% filter(year <= 1998) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1998)
d1999 <- deactivations_county_year %>% filter(year <= 1999) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 1999)
d2000 <- deactivations_county_year %>% filter(year <= 2000) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2000)
d2001 <- deactivations_county_year %>% filter(year <= 2001) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2001)
d2002 <- deactivations_county_year %>% filter(year <= 2002) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2002)
d2003 <- deactivations_county_year %>% filter(year <= 2003) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2003)

d2004 <- deactivations_county_year %>% filter(year <= 2004) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2004)
d2005 <- deactivations_county_year %>% filter(year <= 2005) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2005)
d2006 <- deactivations_county_year %>% filter(year <= 2006) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2006)
d2007 <- deactivations_county_year %>% filter(year <= 2007) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2007)
d2008 <- deactivations_county_year %>% filter(year <= 2008) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2008)
d2009 <- deactivations_county_year %>% filter(year <= 2009) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2009)
d2010 <- deactivations_county_year %>% filter(year <= 2010) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2010)
d2011 <- deactivations_county_year %>% filter(year <= 2011) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2011)
d2011 <- deactivations_county_year %>% filter(year <= 2011) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2011)
d2012 <- deactivations_county_year %>% filter(year <= 2012) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2012)
d2013 <- deactivations_county_year %>% filter(year <= 2013) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2013)

d2014 <- deactivations_county_year %>% filter(year <= 2014) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2014)
d2015 <- deactivations_county_year %>% filter(year <= 2015) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2015)
d2016 <- deactivations_county_year %>% filter(year <= 2016) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2016)
d2017 <- deactivations_county_year %>% filter(year <= 2017) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2017)
d2018 <- deactivations_county_year %>% filter(year <= 2018) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2018)
d2019 <- deactivations_county_year %>% filter(year <= 2019) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2019)
d2020 <- deactivations_county_year %>% filter(year <= 2020) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2020)
d2021 <- deactivations_county_year %>% filter(year <= 2021) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2021)
d2022 <- deactivations_county_year %>% filter(year <= 2022) %>% group_by(FIPS_CODE) %>% count() %>% mutate(year = 2022)

#merge this and then subtract county total-cumulative deactivations
deactivate <- rbind(d1974, d1975, d1976, d1977, d1978, d1979, d1980, d1981, d1982, d1983, d1984, d1985,
                d1986, d1987, d1988, d1989, d1990, d1991, d1992, d1993, d1994, d1995, d1996, d1997,
                d1998, d1999, d2000, d2001, d2002, d2003, d2004, d2005, d2006, d2007, d2008, d2009,
                d2010, d2011, d2012, d2013, d2014, d2015, d2016, d2017, d2018, d2019, d2020, d2021,
                d2022)

deactivate_joined <- left_join(Data, deactivate, by = c("year" = "year", "FIPS" = "FIPS_CODE"))
deactivate_joined$n[is.na(deactivate_joined$n)] <- 0
Data$CumulativeDeactivatedFacilities <- deactivate_joined$n

###### FINALLY, just subtract cumulative deactivated from total and we get the number active per year
Data$ActiveFacilities <- Data$AllFacilities - Data$CumulativeDeactivatedFacilities

######## And we get out primary variable here, # violations / num facilities
Data$NumViolators_NumFacilities <- Data$NumViolatingFacilities / Data$ActiveFacilities
Data$NumViolators_NumFacilities[which(Data$NumViolators_NumFacilities == Inf)] <- 0 #Inf occurs if denominator is zero. I treat this as still a zero.
Data$NumViolators_NumFacilities[is.na(Data$NumViolators_NumFacilities)] <- 0 #there were none, but just in case.
```

```{r makehistogram}
hist(Data$NumViolators_NumFacilities)
mean(Data$NumViolators_NumFacilities)
sd(Data$NumViolators_NumFacilities)
```

## Add in some other control variables to Data

```{r othervariables}
# Population Served
population_sum <- pws_frs %>% group_by(FIPS_CODE) %>% 
  summarise(pop = sum(POPULATION_SERVED_COUNT))

Data$PopulationServed <- rep(as.numeric(population_sum$pop), 49)

# Deactivations per year by county
Deactivated <- left_join(Data, deactivations_county_year, by = c("year" = "year", "FIPS" = "FIPS_CODE"))
Deactivated$n[which(is.na(Deactivated$n))] <- 0

Data$DeactivatedFacilities <- Deactivated$n
```

## Save Data

```{r savedata}
save(Data, file = "/Users/elizabeth/Documents/Capstone Paper/SDWA_Data.Rdata")
```

